### 2018-12-14 writing/thesis/chapters/methods.tex
Improve methods chapter slightly.

% Names of methods or so
\newcommand{\muscl}{\textsc{muscl}-Hancock}
\newcommand{\dg}{\textsc{dg}}
\newcommand{\ader}{\textsc{ader}}
\newcommand{\aderdg}{\textsc{ader-dg}}
\newcommand{\amr}{\textsc{amr}}

@@

\section{Conservative Form}
Standard (hyperbolic) form of conservation law
\begin{equation}
  \label{eq:conservation-law}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q) = \source(\bm{x}, t, \Q)
\end{equation}
extend to:
\begin{equation}
  \label{eq:conservation-law-gradient}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q, \gradQ) = \source(\bm{x}, t, \Q)
\end{equation}

\section{The ADER-DG Method}\label{sec:ader-dg}
We describe the arbitrary derivative discontinous Galerkin (ADER-DG) method in this chapter.
See: \cite{dumbser2018efficient}

We discuss the solution of a hyperbolic conservation law \cref{eq:conservation-law} with domain $\domain$ and boundary $boundary$.
In the discontinous Galerkin (DG) framework, we approximate this solution in the space
\begin{equation}
  \label{eq:dg-space}
  \broken = \bigcup_i \cell
\end{equation}
of disjoint quadrilateral cells $\cell$.
Note that we do not distinguish between the approximation space and the domain, the use should be clear given its surrounding context.
In the following we make use of the Einstein summation convention where summation over repeated indexes is implied.

@@

We now describe the derivation of the \textsc{ader-dg} method.
This scheme is a predictor-corrector method.
We first compute a local solution of the cell in the predictor step and then connect with the neighbors in the corrector step.
In the following we first describe the corrector step as it follows directly from the \textsc{pde}.
The predictor is derived shortly after.

@@

\sidetitle{Corrector}
First, we multiply the system \cref{eq:conservation-law} by a test function $\testfunction{i}$ and integrate over the space-time volume $(\cell \times [t^n, t^{n+1}])$.
We arrive at the so called weak formulation of the \textsc{pde} 
\begin{equation}
  \label{eq:weak-pde}
\intdt{\intdcell{
\testfunction{k} \pdv{\Q}{t}
}}
+
\intdt{\intdcell{
    \testfunction{k} \left( \div{F(\Q, \gradQ} \right)
}}
=
\intdt{\intdcell{
    \testfunction{k} S(\Q, \bm{x}, t)
}}
\end{equation}
\todo{Include gradient everywhere!}
\todo{Double check the following description}
We now replace the solution $\bm{Q}$ with so called spacetime-predictor $\stpredictor (\bm{x},t)$ into the weak form and write it as a polynomial using the representation of \cref{eq:cell-approx}.
Integrate first by parts in time (note basis here not defined over time), flux divergence by parts in space.
\begin{align}
\begin{split}
\label{eq:corrector}
\left(
\intdcell{
  \testfunction{k} \sbasis{l}
}
\right)
(\bm{u^{n+1} - u^{n}})
&+
\left(\intdt{\intdcell{
      \testfunction{k} \Riemann(\stpredictor^-, \stpredictor^+) \cdot \normal
}}\right)
-\\
\left(\intdt{\intdcell{
    \gradient{\testfunction{k}} \cdot  \flux(\stpredictor)
}}\right)
&=
\left(\intdt{\intdcell{
      \testfunction{k} \source(\stpredictor)
}}\right)
\end{split}
\end{align}
The first term motivates an orthogonal basis as this leads to a diagonal mass matrix.

\sidetitle{Predictor}
To derive the predictor, we again take an approximation of our solution in a nodal basis, but now consider polynomials that are defined both in space and time
\begin{equation}
  \label{eq:cell-approx-space-time}
  q(\bm{x}, t^n)_{|\cell[i]} = \hat{u}^n_{i,l} \stbasis{l}(\bm{x}, t).
\end{equation}
We now multiply the conservation law again by a test function (of the same function space as the basis) and arrive at the weak formulation
\begin{equation}\label{eq:weak-pde-space-time}
\intdt{\intdcell{
    \testfunction{k}(\bm{x}, t)
    \pdv{\stpredictor}{t}
}}
+
\intdt{\intdcell{
    \testfunction{k}(\bm{x}, t)
    \left(
      \divergence{\flux(\stpredictor)}
    \right)
}}
=
\intdt{\intdcell{
  \testfunction{k} \source(\stpredictor)
}}.
\end{equation}
Similar to the derivation of the corrector, we again integrate the first term by parts in time and the flux divergence in space.
This time we do not use the Riemann solver for the flux boundary term but rather use the discrete solution at time $t$.
Note that this neglects the interaction with neighbouring cells; this is corrected in the corrector step.
\todo{Replace $\bm{x}$ with a macro some other letter. Is also coordinate!}

\begin{align}\label{eq:space-time-predictor}
\begin{split}
\intdcell{
  \testfunction{k} (\bm{x}, t^{n+1}) \stpredictor(\bm{x}. t^{n+1})
}
&-
\intdt{\intdcell{
    \pdv{}{d} \testfunction{k}(\bm{x}, t) \stpredictor(\bm{x}, t)
}}
-
\intdcell{
  \testfunction{k}(\bm{x},t^n) \dsol(\bm{x}, t^n)
} = \\
\intdt{\intdcell{
    \testfunction{k}(\bm{x},t) \divergence{\flux(\stpredictor)}
}}
&+
\intdt{\intdcell{
    \testfunction{k}(\bm{x}, t) \source(\stpredictor)
}}
\end{split}
\end{align}
Inserting \cref{eq:cell-approx-space-time} results in a local systems of equations that can be solved in a fixed point iteration scheme.
\todo{Inital guess correct? Maybe cite sth?}
As initial value for the space-time-predictor we use the solution of the timestep before.
\todo{Maybe collect into matrices and show scheme?}
For details and proof of convergence for the linear case, see~\cite{dumbser2008unified}.

@@

This Riemann solver was first published in~\cite{gassner2008discontinuous} and used for an \textsc{ader-dg} scheme in~\cite{dumbser2010arbitrary}.
The timestep is restriced to a so-called \textsc{cfl}-type penalty
\begin{equation}
 \Delta t \leq  \text{CFL} \, \frac{\alpha(N) \, h}{\maxConvEigen + 2 \maxViscEigen \frac{2N+1}{h}}
\end{equation}
with $N$ polynomial order and $h$ characteristic length scale of elements~\cite{dumbser2010arbitrary,gassner2008discontinuous}.
The constant $\alpha(N) \leq {\left( 2N+1  \right)}^{-1}$ is obtained from von Neumann analysis on a simple model problem and depends on the approximation order~\cite{dumbser2008unified}.
We use a value of $0.7$ for the constant $\text{CFL}$.
\todo{Describe CFL-cond.\ for multiple dimensions!}

@@

@@

Putting everything together, our complete \textsc{pde} is a conservation law of the form of \cref{eq:conservation-law-gradient}
\begin{equation}
 \begin{array}{l}
 \text{mass cons.} \\
 \text{momentum cons.} \\
 \text{energy cons.} \\
 \text{cont.\ gas} 
\end{array}
:
\quad
  \pdv{}{t}
  \underbrace{
  \begin{pmatrix}
    \Qrho\\
    \Qj\\
    \QE\\
    \QZ
    \end{pmatrix}}_{\Q}
  + 
  \divergence{
  \left(
   \underbrace{\hyperFluxDef}_{\hyperFlux(\Q)}
+
\underbrace{\viscFluxDef}_{\viscFlux(\Q, \gradQ)}
  \right)}
 =
  \underbrace{
  \begin{pmatrix}
    S_{\Qrho\phantom{\Qrho}}\\
    S_{\Qj}\\
    S_{\QE}\\
    S_{\QZ}
    \end{pmatrix}}_{\source(\Q, \bm{x}, t)}
\end{equation}

@@

A relatively recent way of dealing with this problem is the finite volume subcell limiter.
Our discussion follows the one in~\cite{dumbser2016simple}.
This limiting is an a posteriori limiting, which means that we first evaluate a timestep with an unlimited \textsc{ader-dg} method and then check, whether our solution is \enquote{correct}.
In our case, we check whether the solution is a valid floating point number.
We also check whether it is physically admissible.
In our case we use the criterion
\begin{equation}
  \label{eq:limiting-physical}
  \operatorname{is-admissible}(\Q) =
  \begin{cases}
    \text{true} & \Qrho > 0 \land \pressure > 0 \land \QZ > 0\\
    \text{false} & \text{otherwise}.
  \end{cases}
\end{equation}
The positivity of the pressure implies positivity of energy.
\todo{Mention dmp if I end up using it.}

>>>>

% Names of methods or so
\newcommand{\muscl}{\textsc{muscl}-Hancock}
\newcommand{\dg}{\textsc{dg}}
\newcommand{\ader}{\textsc{ader}}
\newcommand{\aderdg}{\textsc{ader-dg}}
\newcommand{\amr}{\textsc{amr}}
\newcommand{\pde}{\textsc{pde}}

@@

\section{The ADER-DG Method}\label{sec:ader-dg}
We describe the arbitrary derivative discontinous Galerkin (\aderdg) method in this chapter.
Our description of the main method follows~\cite{dumbser2008unified,dumbser2010arbitrary,dumbser2018efficient}, our notation follows primarily~\cite{dumbser2018efficient}.
We discuss the approximation of systems of partial differential equations (\pde) of the form
\begin{equation}
  \label{eq:conservation-law-gradient}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q, \gradQ) = \source(\bm{x}, t, \Q),
\end{equation}
which in contrast to hyperbolic conservation laws of the form
\begin{equation}
  \label{eq:conservation-law}
 \frac{\partial}{\partial_t}  \Q + \div \flux(\Q) = \source(\bm{x}, t, \Q)
\end{equation}
contain the gradient $\gradQ$ of the so called vector of conservative variables $\Q$~\cite{dumbser2010arbitrary}.
There are therefore not hyperbolic but rather parabolic or elliptic.

We discuss the solution of a hyperbolic conservation law \cref{eq:conservation-law} with domain $\domain$ and boundary $\boundary$.
Let $\domain$ and $\boundary$ denote our two- or three-dimensional computational domain and its boundary.
In our implementation of the discontinous Galerkin (\dg) framework, we approximate this solution in the space
\begin{equation}
  \label{eq:dg-space}
  \broken = \bigcup_i \cell
\end{equation}
of disjoint quadrilateral cells $\cell$.
Note that we do not distinguish between the broken approximation space and the domain, the use should be clear given its surrounding context.
In the following we make use of the Einstein summation convention where summation over repeated indexes is implied.

@@

We now describe the derivation of the \textsc{ader-dg} method.
This scheme is a predictor-corrector method.
We first compute a local solution of the cell in the predictor step and then connect with the neighbors in the corrector step.
In the following we first describe the corrector step as it follows directly from the \pde.
The predictor is derived shortly after.

@@

\sidetitle{Corrector}
First, we multiply the system \cref{eq:conservation-law} by a test function $\testfunction{i}$ and integrate over the space-time volume $(\cell \times [t^n, t^{n+1}])$.
We arrive at the so called weak formulation of the \pde\
\begin{equation}
  \label{eq:weak-pde}
\intdt{\intdcell{
\testfunction{k} \pdv{\Q}{t}
}}
+
\intdt{\intdcell{
    \testfunction{k} \left( \div{F(\Q, \gradQ} \right)
}}
=
\intdt{\intdcell{
    \testfunction{k} S(\Q, \bm{x}, t)
}}
\end{equation}
\todo{Include gradient everywhere!}
\todo{Double check the following description}
We now replace the solution $\bm{Q}$ with so called spacetime-predictor $\stpredictor (\bm{x},t)$ into the weak form and write it as a polynomial using the representation of \cref{eq:cell-approx}.
Integrate first by parts in time (note basis here not defined over time), flux divergence by parts in space.
\begin{align}
\begin{split}
\label{eq:corrector}
\left(
\intdcell{
  \testfunction{k} \sbasis{l}
}
\right)
(\bm{u^{n+1} - u^{n}})
&+
\left(\intdt{\intdcell{
      \testfunction{k} \Riemann(\stpredictor^-, \stpredictor^+) \cdot \normal
}}\right)
-\\
\left(\intdt{\intdcell{
    \gradient{\testfunction{k}} \cdot  \flux(\stpredictor, \gradient{\stpredictor}) % todo correct gradient?
}}\right)
&=
\left(\intdt{\intdcell{
      \testfunction{k} \source(\stpredictor)
}}\right)
\end{split}
\end{align}
The first term motivates an orthogonal basis as this leads to a diagonal mass matrix.

\sidetitle{Predictor}
To derive the predictor, we again take an approximation of our solution in a nodal basis, but now consider polynomials that are defined both in space and time
\begin{equation}
  \label{eq:cell-approx-space-time}
  q(\bm{x}, t^n)_{|\cell[i]} = \hat{u}^n_{i,l} \stbasis{l}(\bm{x}, t).
\end{equation}
We now multiply the conservation law again by a test function (of the same function space as the basis) and arrive at the weak formulation
\begin{equation}\label{eq:weak-pde-space-time}
\intdt{\intdcell{
    \testfunction{k}(\bm{x}, t)
    \pdv{\stpredictor}{t}
}}
+
\intdt{\intdcell{
    \testfunction{k}(\bm{x}, t)
    \left(
      \divergence{\flux(\stpredictor, \gradient{\stpredictor})}
    \right)
}}
=
\intdt{\intdcell{
  \testfunction{k} \source(\stpredictor)
}}.
\end{equation}
Similar to the derivation of the corrector, we again integrate the first term by parts in time and the flux divergence in space.
This time we do not use the Riemann solver for the flux boundary term but rather use the discrete solution at time $t$.
Note that this neglects the interaction with neighbouring cells; we account for this in the corrector step.
\todo{Replace $\bm{x}$ with a macro some other letter. Is also coordinate!}

\begin{align}\label{eq:space-time-predictor}
\begin{split}
\intdcell{
  \testfunction{k} (\bm{x}, t^{n+1}) \stpredictor(\bm{x}, t^{n+1})
}
&-
\intdt{\intdcell{
    \pdv{}{t} \testfunction{k}(\bm{x}, t) \stpredictor(\bm{x}, t)
}}
-
\intdcell{
  \testfunction{k}(\bm{x},t^n) \dsol(\bm{x}, t^n)
} = \\
\intdt{\intdcell{
    \testfunction{k}(\bm{x},t) \divergence{\flux(\stpredictor, \gradient{\stpredictor})}
}}
&+
\intdt{\intdcell{
    \testfunction{k}(\bm{x}, t) \source(\stpredictor)
}}
\end{split}
\end{align}
Inserting \cref{eq:cell-approx-space-time} results in a local systems of equations that can be solved in a fixed point iteration scheme.
\todo{Inital guess correct? Maybe cite sth?}
As initial value for the space-time-predictor we use the solution of the timestep before.
\todo{Maybe collect into matrices and show scheme?}
For details and proof of convergence for the linear case, see~\cite{dumbser2008unified}.

@@

This Riemann solver was first published in~\cite{gassner2008discontinuous} and used for an \textsc{ader-dg} scheme in~\cite{dumbser2010arbitrary}.
The timestep is restriced to a so-called \textsc{cfl}-type penalty
\begin{equation}\label{eq:cfl-aderdg}
 \Delta t \leq  \text{CFL} \, \frac{\alpha(N) \, h}{\maxConvEigen + 2 \maxViscEigen \frac{2N+1}{h}}
\end{equation}
with $N$ polynomial order and $h$ characteristic length scale of elements~\cite{dumbser2010arbitrary,gassner2008discontinuous}.
The constant $\alpha(N) \leq {\left( 2N+1  \right)}^{-1}$ is obtained from von Neumann analysis on a simple model problem and depends on the approximation order~\cite{dumbser2008unified}.
We use a value of $0.7$ for the constant $\text{CFL}$.
\todo{Describe CFL-cond.\ for multiple dimensions!}

@@

We again use the Rusanov-type Riemann solver~\cref{eq:rusanov-flux}.
The timestep is subject to a penalty of the form
\begin{equation}\label{eq:cfl-muscl}
 \Delta t \leq  \text{CFL} \, \frac{h}{\maxConvEigen + 2 \maxViscEigen \frac{1}{h}},
\end{equation}
with a constant value of CFL that is typically close to $0.7$.

@@

Putting everything together, our complete \pde\ is a conservation law of the form of \cref{eq:conservation-law-gradient}
\begin{equation}
 \begin{array}{l}
 \text{mass cons.} \\
 \text{momentum cons.} \\
 \text{energy cons.} \\
 \text{cont.\ gas} 
\end{array}
:
\quad
  \pdv{}{t}
  \underbrace{
  \begin{pmatrix}
    \Qrho\\
    \Qj\\
    \QE\\
    \QZ
    \end{pmatrix}}_{\Q}
  + 
  \divergence{
  \left(
   \underbrace{\hyperFluxDef}_{\hyperFlux(\Q)}
+
\underbrace{\viscFluxDef}_{\viscFlux(\Q, \gradQ)}
  \right)}
 =
  \underbrace{
  \begin{pmatrix}
    S_{\Qrho\phantom{\Qrho}}\\
    S_{\Qj}\\
    S_{\QE}\\
    S_{\QZ}
    \end{pmatrix}}_{\source(\Q, \bm{x}, t)}
\end{equation}

@@

A relatively recent way of dealing with this problem is the finite volume subcell limiter.
Our discussion follows the one in~\cite{dumbser2016simple}.
This limiting is an a posteriori limiting, which means that we first evaluate a timestep with an unlimited \textsc{ader-dg} method and then check, whether our solution is \enquote{correct}.
In our case, we check whether the solution is a valid floating point number.
We also check whether it is physically admissible.
In our case we use the criterion
\begin{equation}
  \label{eq:limiting-physical}
  \operatorname{is-admissible}(\Q) =
  \begin{cases}
    \text{true} & \Qrho > 0 \land \pressure > 0 \land \left(0 \leq \QZ \leq 1 \right)\\
    \text{false} & \text{otherwise}.
  \end{cases}
\end{equation}
The positivity of the pressure implies positivity of energy.
\todo{Mention dmp if I end up using it.}

If a solution violates our criteria, we subdivide the infringing cell into $N_2 = 2N + 1$ subcells.
This results in the remarkable properties, that we first do not loose any spatial resolution and secondly that we can use the same timestep size for both methods.
This can be seen easily by noticing the similar form of \cref{eq:cfl-aderdg} and \cref{eq:cfl-muscl}.
